---
title: "Fast Rmarkdown Theming with `thematic` and `bootstraplib`"
author: "Thomas Vroylandt"
date: "04/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
library(DiagrammeR)
library(here)
library(knitr)
library(tidyverse)
library(kableExtra)
```

Theming in Rmarkdown can be hard. You first made some custom CSS or use a provided theme but your figures didn't change and you have to style your `ggplot2` theme. And after it you change your mind (or your boss do) and you need to move this color shade to lighter one. So you change your theme, but forgot to change it in all your CSS and something is going wrong. Same goes for fonts...

Sounds familiar ? Theming is a common problem, because it involves lot of differents elements which aren't ruled by the same command. Your process look like this.

<center>
```{r, fig.height=2}
grViz("
digraph {

  graph [overlap = true, fontsize = 10, rankdir = LR]

  node [shape = oval,
        fontname = Lato,
        color = '#000000',
        fontcolor = '#000000']
    rmd [label = 'RMarkdown']
    ggplot [label = 'gglot2 theme']
    css [label = 'Custom CSS']
    output [label = 'Output document']

  edge [arrowhead = vee,
        arrowsize = 0.5,
        color = '#000000']
    rmd -> {ggplot css}
    {ggplot css} -> output
  }
      ")
```
</center>

Thanks to the new `thematic` package and the not so old `bootstraplib` package, you can fastly build themes for your documents without the problems mentioned above.

# Style your graph theme with `thematic`

<center>
```{r, fig.height=2, fig.align='center'}
grViz("
digraph {

  graph [overlap = true, fontsize = 10, rankdir = LR]

  node [shape = oval,
        fontname = Lato,
        color = '#000000',
        fontcolor = '#000000']
    rmd [label = 'RMarkdown']
    thematic [label = 'thematic rules', 
              color = '#741336', 
              fontcolor = '#741336',
              penwidth = 2.5]
    css [label = 'Custom CSS']
    output [label = 'Output document']

  edge [arrowhead = vee,
        arrowsize = 0.5,
        color = '#000000']
    rmd -> css
    rmd -> thematic [color = '#741336']
    thematic -> output [color = '#741336']
    css -> output
  }
      ")
```
</center>

If you want to read the docs :

+ https://rstudio.github.io/thematic/ ;
+ https://bit.ly/thematic-talk.

## How to use `thematic`

You can style six differents elements :

+ `bg` : the background color, here <b style="background-color:#F4F4E2">#F4F4E2</b> ;
+ `fg` : the foreground color, here <b style="background-color:#4c4c4c;color:white">#4c4c4c</b> ;
+ `accent` : the accent color, here <b style="background-color:#741336;color:white">#741336</b>. It will also be used to create a sequential palette ;
+ `font` : the font you want to use, here <b style="font-family:'Special Elite'">Special Elite</b>. You can go further by using `font_spec()` ;
+ `qualitative` : the qualitative palette you want to use ;
+ `sequential` : the sequential palette, which can be controlled using `sequential_gradient()`. By default, it creates a palette from a mix between your background color and your accent color to a mix between your foreground color and your accent color.

You just have to use `thematic_on()` with your parameters to use it.

```{r, echo = TRUE}
library(thematic)
library(ggplot2)
library(tibble)

thematic_on(
  bg = "#F4F4E2",
  fg = "#4c4c4c",
  accent = "#741336",
  font = font_spec("Special Elite", scale = 0.5),
  qualitative = paletteer::paletteer_d("dutchmasters::milkmaid"),
  sequential = sequential_gradient(0.5, 0.75)
)
```

After that, just make a normal plot like you are used to.

```{r, echo = TRUE, fig.align='center'}
ggplot(tibble(mtcars,
              model = rownames(mtcars)),
       aes(x = model, y = mpg, fill = disp)) +
  geom_col() +
  geom_text(aes(label = disp), 
            position = position_stack(vjust = 1.1)) +
  coord_flip() +
  labs(title = "A barplot with mtcars",
       caption = "Source : mtcars data")
```

You can add a theme to your graph but will need some further coding. The last element overlap your styling (which is a good thing) but because complete themes have fonts and/or colors, they may replace your styling, which is bad.

```{r, echo = TRUE, fig.align='center'}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color =  Species)) +
  geom_point(size = 1.1) +
  geom_smooth() +
  labs(title = "A nice scatterplot for iris",
       caption = "Source : Iris data") +
  theme_minimal(base_family = thematic_get_option("font")$families,
                base_size = thematic_get_option("font")$scale * 11,
                base_line_size = 1.1)
```

```{r}
thematic_off()
```

## Some improvements

This is a good package, but it's only the 0.0.1 version and some improvements could be done.

+ For the moment, `thematic` doesn't work with interactive graph packages like `plotly` or `ggiraph`. But maybe in the future (see https://talks.cpsievert.me/20200429/#24) ?

+ Using completes themes get complicated because it removes some of the choices you did. So you need more styling through the `theme` command to makes everything fine after it. Maybe it would be useful to adapt some themes to make them compatible by default with `thematic` (see https://github.com/rstudio/thematic/issues/16) ? 

# Use `bootstraplib` for Rmarkdown theming

<center>
```{r, fig.height=2, fig.align='center'}
grViz("
digraph {

  graph [overlap = true, fontsize = 10, rankdir = LR]

  node [shape = oval,
        fontname = Lato,
        color = '#000000',
        fontcolor = '#000000']
    rmd [label = 'RMarkdown']
    ggplot [label = 'gglot2 theme']
    bootstraplib [label = 'bootstraplib', 
                  color = '#741336', 
                  fontcolor = '#741336',
                  penwidth = 2.5]
    output [label = 'Output document']

  edge [arrowhead = vee,
        arrowsize = 0.5,
        color = '#000000']
    rmd -> ggplot 
    rmd -> bootstraplib [color = '#741336']
    bootstraplib -> output [color = '#741336']
    ggplot -> output
  }
      ")
```
</center>

Some ressources on the subject :

+ https://rstudio.github.io/bootstraplib/ ;
+ https://resources.rstudio.com/rstudio-conf-2020/styling-shiny-apps-with-sass-and-bootstrap-4-joe-cheng?prevItm=0&prevCol=6275649&ts=6836.

## How to use `bootstraplib`

`bootstraplib` calls Bootstrap and allow you to style directly your document (RMarkdown or Shiny) from R using Sass. You can do a lot with this package. But we will only cover some part of it.

First you need to reference the use of Bootstrap in the YAML header of your Rmd.

```{r, eval = FALSE, echo = TRUE}
---
output:
  html_document:
    bootstrap_version: 4+3
---
```

After that you can use some functions to style your main elements :

+ `bs_theme_base_colors()` allows you to set you background (with `bg`) and foreground (with `fg`) colors ;
+ `bs_theme_accent_colors()` for the accent colors ;
+ `bs_theme_fonts()` for the fonts.

You can also add some hand-made rules with `bs_theme_add_variables()`. As an example `bs_theme_add_variables("font-size-base" = "0.9rem")` will style the size of your fonts.

If you want to use Google Fonts, you may need the `gfonts` package, made by dreamRs (https://github.com/dreamRs/gfonts). It offers simple functions to deal with Google Fonts.

As an example, here is the code chunk which creates the document just below.

```{r, eval=FALSE, echo = TRUE}
# bootstraplib part
library(bootstraplib)
library(gfonts)

# setup and use font
# the setup has to be done once
setup_font("special-elite", "www/")
use_font("special-elite", "www/css/special-elite.css")

bs_theme_fonts(base = "Special Elite")

# font-size
bs_theme_add_variables("font-size-base" = "0.9rem")

# colors
bs_theme_base_colors(bg = "#F4F4E2",
                     fg = "#4c4c4c")

bs_theme_accent_colors(primary = "#741336")
```

<iframe 
  width = "100%"
  src ="examples/bootstraplib_gfonts.html">
</iframe>

## Improvements

+ The package will need to deal with more outputs. It doesn't work with `xaringan` and `pagedown` for the moment and strange with `flexdashboard`.

+ You will need to style your table manually. For the moment, only `kable` works well with `bootstraplib`. The other packages I tested all failed on one part :

  + `gt` has the font family only ;
  + `reactable` has the font size, family and color ;
  + `flextable` has the right background color but not the font size/family/color.

# Both of them is even more powerful

<center>
```{r, fig.height=1.5, fig.width = 8, fig.align='center'}
grViz("
digraph {

  graph [overlap = true, fontsize = 10, rankdir = LR]

  node [shape = oval,
        fontname = Lato,
        color = '#000000',
        fontcolor = '#000000']
    rmd [label = 'RMarkdown']
    thematic [label = 'thematic rules', 
              color = '#741336', 
              fontcolor = '#741336',
              penwidth = 2.5]
    bootstraplib [label = 'bootstraplib', 
                  color = '#741336', 
                  fontcolor = '#741336',
                  penwidth = 2.5]
    output [label = 'Output document']

  edge [arrowhead = vee,
        arrowsize = 0.5,
        color = '#741336']
    rmd -> bootstraplib -> thematic -> output
  }
      ")
```
</center>

If you use `thematic` after `bootstraplib` in a Rmarkdown document (it's also true for Shiny), it will retrieve the styling you passed and use them. That's a very powerful feature because you will only need to style your document once.

The only thing you need to care about is the font. You have to set it to `auto` inside `thematic_on()` to get it nicely.

Here is the corresponding functions between the two packages :

```{r}
tibble(
  name = c(
    "Background color",
    "Foreground color",
    "Accent color",
    "Font family",
    "Font size",
    "Sequential palette",
    "Qualitative palette"
  ),
  thematic = c(
    "`bg`",
    "`fg`",
    "`accent`",
    "`font`",
    "`font_spec(scale = )`",
    "`sequential`",
    "`qualitative`"
  ),
  bootstraplib = c(
    "`bs_theme_base_colors(bg = )`",
    "`bs_theme_base_colors(fg = )`",
    "`bs_theme_accent_colors(primary = )`",
    "`bs_theme_fonts(base = )`",
    "`bs_theme_add_variables('font-size-base' = )`",
    "",
    ""
  ),
  both = c(
    "auto",
    "auto",
    "auto",
    "`font_spec('auto')`",
    "`font_spec('auto', scale = )`",
    "auto",
    ""
  ),
  info = c(
    "",
    "",
    "",
    "May need `gfonts`",
    "",
    "Calculated from background, foreground and accent color",
    "Manual"
  )
) %>%
  kable(
    "html",
    col.names = c(
      "Feature",
      "Using `thematic`",
      "Using `bootstraplib`",
      "Using both",
      "Additional infos"
    )
  ) %>%
  kable_styling()
```

As an example, if you use both packages together.

```{r, eval=FALSE, echo = TRUE}
# bootstraplib part
library(bootstraplib)
library(gfonts)
library(thematic)

# setup and use fonts - google fonts
# the setup has to be done once
setup_font("special-elite", "www/")
use_font("special-elite", "www/css/special-elite.css")
bs_theme_fonts(base = "Special Elite")

# font-size
bs_theme_add_variables("font-size-base" = "0.9rem")

# colors
bs_theme_base_colors(bg = "#F4F4E2",
                     fg = "#4c4c4c")

bs_theme_accent_colors(primary = "#741336")

# thematic part
thematic_on(font = font_spec("auto", scale = 0.5))
```

<iframe 
  width = "100%"
  src ="examples/bootstraplib_thematic.html">
</iframe>

# Conclusion

This two packages don't do everything but it's a good step if you don't want to spend a lot of time styling your RMarkdown document. It could also be a first step to more personalized styling as you can export it.

Personly, I really like the "control tower" approach. It doesn't force you to repeat your styling and makes it really easier to change.
